---
title: "Quick Start"
author:
  - name: Herdiantri Sufriyana
    affiliation:
    - &ibi Institute of Biomedical Informatics, College of Medicine, National 
      Yang Ming Chiao Tung University, Taipei, Taiwan. 
    email: herdi@nycu.edu.tw
  - name: Emily Chia-Yu Su
    affiliation:
    - *ibi
date: "2024-12-24"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
vignette: >
  %\VignetteIndexEntry{Machine Learning Nomogram Exemplar}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Programming environment

```{r Load necessary packages}
library(rmlnomogram)
```

# Binary outcome (or class-wise multinomial outcome)

## 1 - Categorical predictors and binary outcome without probability

```{r Example 1 - Without explainability, fig.height=3, fig.width=12}
data(nomogram_features)
data(nomogram_outputs)
create_nomogram(nomogram_features, nomogram_outputs)
```

## 2 - Categorical predictors and binary outcome with probability

```{r Example 2 - Without explainability, fig.height=4.5, fig.width=10}
create_nomogram(nomogram_features, nomogram_outputs, prob = TRUE)
```

```{r Example 2 - With explainability, fig.height=6, fig.width=12}
data(nomogram_shaps)
create_nomogram(
  nomogram_features, nomogram_outputs, nomogram_shaps
  , prob = TRUE
)
```

## 3 - Categorical and 1 numerical predictors and binary outcome with probability

```{r Example 3 - Without explainability, fig.height=7.5, fig.width=12}
data(nomogram_features2)
data(nomogram_outputs2)
create_nomogram(nomogram_features2, nomogram_outputs2, prob = TRUE)
```

```{r Example 3 - With explainability, fig.height=7.5, fig.width=12}
data(nomogram_shaps2)
create_nomogram(
  nomogram_features2, nomogram_outputs2, nomogram_shaps2
  , prob = TRUE
)
```

# Continuous outcome

## 4 - Categorical predictors and continuous outcome

```{r Example 4 - Without explainability, fig.height=4.5, fig.width=10}
data(nomogram_features3)
data(nomogram_outputs3)
create_nomogram(nomogram_features3, nomogram_outputs3, est = TRUE)
```

```{r Example 4 - With explainability, fig.height=4.5, fig.width=10}
data(nomogram_shaps3)
create_nomogram(
  nomogram_features3, nomogram_outputs3, nomogram_shaps3
  , est = TRUE
)
```

## 5 - Categorical and 1 numerical predictors and continuous outcome

```{r Example 5 - Without explainability, fig.height=7.5, fig.width=12}
data(nomogram_features4)
data(nomogram_outputs4)
create_nomogram(nomogram_features4, nomogram_outputs4, est = TRUE)
```

```{r Example 5 - With explainability, fig.height=7.5, fig.width=12}
data(nomogram_shaps4)
create_nomogram(
  nomogram_features4, nomogram_outputs4, nomogram_shaps4
  , est = TRUE
)
```

```{r session-info, echo=TRUE}
sessionInfo()
```
